The Form of Uncertainty Affects Selection for Social Learning

### 2023-02-24 ms.tex
updated SL ceiling

Our model was implemented in the Julia programming language~\citep{Bezanson2017} 
using the Agents.jl agent-based modeling library~\citep{Datseris2022} and run
on the Sherlock supercomputing cluster at Stanford University. All Analysis figures
except Figure~\ref{fig:extinctionVariability} were made using the Julia plotting
library Gadfly~\citep{Jones2021b}. Figure~\ref{fig:extinctionVariability} was made in R~\citep{R}
using the ggplot2 package~\citep{ggplot2}. 
Model and analysis code is publicly available on GitHub at \url{https://github.com/mt-digital/UncMod}.

@@

\edit{Across all parameter settings, we observed that there exists some threshold value of environmental
  variability beyond which the evolution of social learning is exceedingly rare.  We call this the
  \emph{social learning extinction variability}, denoted $u_\text{e}$ (Figure~\ref{fig:extinctionVariability}), which is expressed formally as 
\begin{equation}
  u_\text{e} = \min~\{ u~|~\meansl < 0.01 \}.
  \label{eq:extinctionVariability}
\end{equation}
\noindent
In other words, $u_\text{e}$ is the smallest value of $u$ for which the social learning fixated
less than one in one hundred times. We will use this auxiliary measure to analyze the effect of
the other uncertainty parameters and to highlight that social learning frequently 
evolves despite high levels of environmental variability.}

The number of behavioral options, i.e., larger $B$, favors the evolution of social learning. Increasing the selection-set size
expands the range of $u$ over which social learning is favored for two reasons. First, when the environment remains constant between
generations, the value of social learning increases with more behavioral options, because it effectively increases the number of
observations that one can learn from by comparing the payoffs of several models. In other words, relying \emph{only} on one's
individual trial-and-error learning is less likely to yield useful information when the search space is
large. Second, when the environment changes, bias against the new optimal is weaker with more behavioral
options. For example, with only two behavioral choices ($B = 2$), a common assumption in many models, social
learning is particularly detrimental when the environment changes since more of the population will have
arrived at the optimal behavior in the previous generation and socially learning from them will be biased
against the correct choice. These two mechanisms make it particularly profitable to engage in social
learning when the selection-set size is larger---for large values of $B$ social learning can even be favored
when the environment is more likely to change than not (i.e.,
$u>0.5$\edit{---see Figure~\ref{fig:extinctionVariability}}).

@@

\begin{figure}
  \caption{As the selection-set size, $B$, and effective lifespan, $L$, increase (set equal along x-axis), 
  social learning frequently evolves across an increasingly large range of
  environmental variability values, i.e., up to
  increasingly large values of social learning extinction variability
(y-axis; defined in Equation~\ref{eq:extinctionVariability}).}
  \label{fig:extinctionVariability}
  \centering
    \includegraphics[width=0.8\textwidth]{Figures/SL_Extinction_Variability.pdf}
\end{figure}

@@

Our model necessarily made simplifying assumptions. These were chosen judiciously to
address the main question of how uncertainty affects the evolution of social
learning. Two critical assumptions we made have important consequences for comparing
our results with previous findings in the literature. First, we assumed all agents
engage in individual learning. This means that individual learning is effectively
free, though potentially inaccurate. Although producing new knowledge may be costly,
many other adaptive problems take the form of essentially cost-free individual
learning. For example, it is hard not to encode association between clouds and rain,
between thirst and drinking water, or between friendship and warm laughter.
\edit{If there were an exogenous cost of individual learning, this would likely increase the social learning extinction variability across uncertainty parameter settings.
Our model operationalizes environmental variability as fully stochastic,
which forces evolution towards either a population of social or asocial learners
since environmental change is unpredictable, unlike precursor
models that allowed for persistent frequency
dependence~\citep{BoydRicherson1985,Rogers1988,Feldman1996}.} Second, we only allow for
intergenerational social learning and environmental change. Within generations,
agents can only learn individually. This assumption allows us to replicate the
classic finding that if the environment changes too quickly, socially-learned
information may become \edit{outdated. Adding horizontal social learning between
vertical/oblique learning in this model would likely expand the range of
environmental variability over which social learning would
evolve~\citep{Turner2022}, i.e., horizontal learning would increase what we have
called the social learning extinction variability. However, depending on the
setting, there may be little difference between horizontal and oblique learning, for
example if the generations are purely cultural ‘generations’ where parents, and
others of the parent's generation, may also be peers.}

>>>>

Our model was implemented in the Julia programming language~\citep{Bezanson2017} 
using the Agents.jl agent-based modeling library~\citep{Datseris2022} and run
on the Sherlock supercomputing cluster at Stanford University. All Analysis figures
except Figure~\ref{fig:slCeiling} were made using the Julia plotting
library Gadfly~\citep{Jones2021b}. Figure~\ref{fig:slCeiling} was made in R~\citep{R}
using the ggplot2 package~\citep{ggplot2}. 
Model and analysis code is publicly available on GitHub at \url{https://github.com/mt-digital/UncMod}.

@@

\edit{Across all parameter settings, we observed that there exists some threshold value of environmental
  variability beyond which the evolution of social learning is exceedingly rare.  We call this the
  \emph{social learning ceiling}, denoted $u_\text{c}$ (Figure~\ref{fig:slCeiling}), which is expressed formally as 
\begin{equation}
  u_\text{c} = \min~\{ u~|~\meansl < 0.01 \}.
  \label{eq:slCeiling}
\end{equation}
\noindent
In other words, $u_\text{c}$ is the smallest value of $u$ for which social learning fixated
less than one in one hundred times. We will use this auxiliary measure to analyze the effect of
the other uncertainty parameters and to highlight that social learning frequently 
evolves despite high levels of environmental variability.}

The number of behavioral options, i.e., larger $B$, favors the evolution of social learning. Increasing the selection-set size
expands the range of $u$ over which social learning is favored for two reasons. First, when the environment remains constant between
generations, the value of social learning increases with more behavioral options, because it effectively increases the number of
observations that one can learn from by comparing the payoffs of several models. In other words, relying \emph{only} on one's
individual trial-and-error learning is less likely to yield useful information when the search space is
large. Second, when the environment changes, bias against the new optimal is weaker with more behavioral
options. For example, with only two behavioral choices ($B = 2$), a common assumption in many models, social
learning is particularly detrimental when the environment changes since more of the population will have
arrived at the optimal behavior in the previous generation and socially learning from them will be biased
against the correct choice. These two mechanisms make it particularly profitable to engage in social
learning when the selection-set size is larger---for large values of $B$ social learning can even be favored
when the environment is more likely to change than not (i.e.,
$u>0.5$\edit{---see Figure~\ref{fig:slCeiling}}).

@@

\begin{figure}
  \caption{As the selection-set size, $B$, and effective lifespan, $L$, increase (set equal along x-axis), 
  social learning frequently evolves across an increasingly large range of
  environmental variability values, i.e., up to
  increasingly large values of the social learning ceiling
(y-axis; defined in Equation~\ref{eq:slCeiling}).}
  \label{fig:slCeiling}
  \centering
    \includegraphics[width=0.8\textwidth]{Figures/SL_Extinction_Variability.pdf}
\end{figure}

@@

Our model necessarily made simplifying assumptions. These were chosen judiciously to
address the main question of how uncertainty affects the evolution of social
learning. Two critical assumptions we made have important consequences for comparing
our results with previous findings in the literature. First, we assumed all agents
engage in individual learning. This means that individual learning is effectively
free, though potentially inaccurate. Although producing new knowledge may be costly,
many other adaptive problems take the form of essentially cost-free individual
learning. For example, it is hard not to encode association between clouds and rain,
between thirst and drinking water, or between friendship and warm laughter.
\edit{If there were an exogenous cost of individual learning, this would likely
  increase the social learning ceiling across uncertainty parameter settings.
Our model operationalizes environmental variability as fully stochastic,
which forces evolution towards either a population of social or asocial learners
since environmental change is unpredictable, unlike precursor
models that allowed for persistent frequency
dependence~\citep{BoydRicherson1985,Rogers1988,Feldman1996}.} Second, we only allow for
intergenerational social learning and environmental change. Within generations,
agents can only learn individually. This assumption allows us to replicate the
classic finding that if the environment changes too quickly, socially-learned
information may become \edit{outdated. Adding horizontal social learning between
vertical/oblique learning in this model would likely expand the range of
environmental variability over which social learning would
evolve~\citep{Turner2022}, i.e., horizontal learning would increase what we have
called the social learning ceiling. However, depending on the
setting, there may be little difference between horizontal and oblique learning, for
example if the generations are purely cultural ‘generations’ where parents, and
others of the parent's generation, may also be peers.}
